Storm Data Analysis - Week 4

### 2021-03-28 Manuscript.Rmd
Finalization

---
title: "Storm Data Analysis - Week 4"
author: "Marco Baciarello"
date: "3/17/2021"
output: html_document
---

@@

@@

# Initialize list, vector and counter for the loop
old_events <- vector("list", 14)
new_events <- c()
i <- 1

suppressWarnings(
    # For every new event category
    for (i in seq(1:length(new_fcts))) {
        # Prepare storage array
        old_instances <- c()
        k <- 1
        # Loop over elements in list element
        for (k in seq(1:length(new_fcts[[i]]))) {
            # Match old event instances
            old_instances <- c(old_instances,
                               str_subset(sel_data2$event, new_fcts[[i]][k])
                                )
            k <- k + 1
        }
        # Add to list
        old_events[[i]] <- unique(old_instances)
        i <- i + 1
    }
)
# Name (tag) list elements
names(old_events) <- names(new_fcts)

# TODO: switch to using list names instead of manual entry of new factor levels
sel_data2$event2 <- fct_collapse(
  sel_data2$event,
  hail = old_events$hail,
  tornado = old_events$tornado,
  tsunami = old_events$tsunami,
  stormy = old_events$stormy,
  lightning = old_events$lightning,
  flood = old_events$flood,
  cold = old_events$cold,
  heat = old_events$heat,
  windy = old_events$windy,
  maritime = old_events$maritime,
  fog = old_events$fog,
  drought = old_events$drought,
  fire = old_events$fire,
  landslide = old_events$landslide,
  other_level = "Other events"
) %>%
  fct_recode(
    "Hail" = "hail",
    "Winds" = "windy",
    "Tornadoes" = "tornado",
    "Floods" = "flood",
    "Lightnings" = "lightning",
    "Cold weather" = "cold",
    "Storms/Rains" = "stormy",
    "Oceanic events" = "maritime",
    "Fires" = "fire",
    "Droughts" = "drought",
    "Hot weather" = "heat",
    "Fog" = "fog",
    "Landslides" = "landslide",
    "Tsunamis" = "tsunami"
  ) %>%
  # Reorder according to frequency, desc
  fct_infreq()

@@

@@

```{r overall-plot}

# Create plot data, use dtplyr for speed and convenience
pdata <- lazy_dt(sel_data2)

# Plotting total instances of each event category
# pdata %>%
#     group_by(event2) %>%
#     summarise(
#         count = n(),
#         
#     )

@@

>>>>

---
title: "Storm Data Analysis - Week 4"
author: "Marco Baciarello"
date: "3/17/2021"
output: html_document
  fig_caption: yes
---

@@

Nested loops is our way of going through a factor variable and associating each value to a value in a new factor variable (many-to-one relationship.). The need for loops seems to be related to the fact that these vars are stored as factors, and the lack of a specific function in the forcats package; fct_collapse() might have been one solution, but it takes the exact factor level names—since there were almost 1000 different possible levels, it would have been impractical to call this function.

@@

# Initialize list, vector and counter for the loop
old_events <- vector("list", 14)
new_events <- c()
i <- 1
```

We create a list of possible synonims for weather events. In the inner loop, the factor variable is run (using `stringr::subset()` sequentially against each list element and its children. Where there's a match, the function returns the full value of the variable. We aggregate these values in a vector. In the outer loop, the unique values (i.e., factor levels) are appended to a second list. The second list has a list-element for each of the new factor levels; the old factor levels are thus stored as children of the new ones, based on matching with our criteria.

Once complete, the list _could_ be used as a named vector. However, for some reason we had to insert each element manually using list$element syntax.

```{r factor-loops}
suppressWarnings(
    # For every new event category
    for (i in seq(1:length(new_fcts))) {
        # Prepare storage array
        old_instances <- c()
        k <- 1
        # Loop over elements in list element
        for (k in seq(1:length(new_fcts[[i]]))) {
            # Match old event instances
            old_instances <- c(old_instances,
                               str_subset(sel_data2$event, new_fcts[[i]][k])
                                )
            k <- k + 1
        }
        # Add to list
        old_events[[i]] <- unique(old_instances)
        i <- i + 1
    }
)
# Name (tag) list elements
names(old_events) <- names(new_fcts)

# TODO: switch to using list names instead of manual entry of new factor levels
sel_data2$event2 <- fct_collapse(
  sel_data2$event,
  hail = old_events$hail,
  tornado = old_events$tornado,
  tsunami = old_events$tsunami,
  stormy = old_events$stormy,
  lightning = old_events$lightning,
  flood = old_events$flood,
  cold = old_events$cold,
  heat = old_events$heat,
  windy = old_events$windy,
  maritime = old_events$maritime,
  fog = old_events$fog,
  drought = old_events$drought,
  fire = old_events$fire,
  landslide = old_events$landslide,
  other_level = "Other events"
) %>%
  # Adjust spelling/capitalization of factor levels
  fct_recode(
    "Hail" = "hail",
    "Winds" = "windy",
    "Tornadoes" = "tornado",
    "Floods" = "flood",
    "Lightnings" = "lightning",
    "Cold weather" = "cold",
    "Storms/Rains" = "stormy",
    "Oceanic events" = "maritime",
    "Fires" = "fire",
    "Droughts" = "drought",
    "Hot weather" = "heat",
    "Fog" = "fog",
    "Landslides" = "landslide",
    "Tsunamis" = "tsunami"
  ) %>%
  # Reorder according to frequency, desc
  fct_infreq()

@@

```{r figure-captions}
fig.1.caption = "**Fig. 1:** Figure 1"
```

@@

Data on agricultural (crop) and property damages show unequivocal predominance of flooding phenomena as causes. At `r scales::dollar(dmgdata$total_dmg[1])`, overflowing of oceans, lakes and rivers caused about 4 times as much damage as storms and other rain phenomena, a distant second in ranking. One can easily imagine damages from disasters such as hurricanes Katrina and Sandy are the primary drivers of this result.

```{r damage-plot, fig.cap=fig.1.caption}

# Make a copy of data to be sorted by damage total, descending, for use in
# damage-related plots.
dmgdata <- sel_data2 %>%
    group_by(event2) %>%
    summarise(total_dmg = sum(total_dmg),
              total_inj = sum(injuries),
              total_fat = sum(fatalities)
              ) %>%
    mutate(event2 = as.character(event2)) %>%
    arrange(desc(total_dmg)) %>%
    as_tibble()

### Damage plot

dmgdata %>%
  ggplot(aes(
    x = reorder(event2, -total_dmg),
    y = total_dmg)) +
  geom_col() +
  scale_y_continuous(
    labels = scales::label_dollar(prefix = NULL,
                                  scale = 1e-9,
                                  largest_with_cents = 1),
    name = "Total estimated damages caused\n(in billions of dollars)\n"
  ) + 
  scale_x_discrete("\nWeather event category (1950–2013)") +
  coord_cartesian(ylim = c(0, 5e10)) +
  annotate("segment", x= 2.2, xend = 1.5, y = 50e9, yend = 52e9,
           arrow = arrow(angle = 20,
                         length = unit(0.15, "inches"),
                         type = "closed")
  ) + 
  annotate("text",
           x = 4.3, y = 49e9,
           label = "Plot clipped for readability.\nActual value: $147 billion") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1))
#####

@@

```{r injuries-plot}
### Injuries plot

dmgdata %>%
    ggplot(aes(
        x = reorder(event2, -total_inj),
        y = total_inj)) +
    geom_col() +
    scale_x_discrete("\nWeather event category (1950–2013)") +
    scale_y_continuous(name = "Related injuries reported\n",
                       breaks = seq(0, 1e5, 1e4),
                       labels = scales::label_comma()) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 70, hjust = 1))

```{r fatalities-plot}
### Fatalities plot

dmgdata %>%
    ggplot(aes(
        x = reorder(event2, -total_fat),
        y = total_fat)) +
    geom_col() +
    scale_x_discrete("\nWeather event category (1950–2013)") +
    scale_y_continuous(name = "Related fatalities reported\n",
                       #breaks = scales::extended_breaks(),
                       breaks = seq(0, 5e3, 1e3),
                       labels = scales::label_comma()) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 70, hjust = 1))
```

## Conclusions
